Quantifying unmet ICU beds in tertiary hospitals in urban India using Propensity

### 2022-09-23 manuscript.Rmd
Update manuscript.Rmd

Trauma globally and in India
----------------------------
Trauma is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC. Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma. The annual mortality from injury in India amount to one million casualties @onemilliondead. Half of trauma mortality in India occur in-hospital, and in-hospital trauma mortality in Indian hospitals was double that of high-income countries @indiantrauma.

@@

Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but are not because such resources are scarce, overloaded and unavailable. Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay. Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @livessaved

A diverse mix of public, for-profit and not-for-profit private actors provide health services in India. According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds. This is 8.2 per 100 000 inhabitants.
The WHO report also highlight the probable shortage of ICU resources in India, though in the context of the COVID-19 pandemic.

@@

One aspect of TQIP assessments is to collect trauma data from participating health care providers and apply statistical analysis to the data. From this data, TQIPs utilizes a retrospective cohort study, analysing the current of trauma care patients and post-fact attempting to determine the appropriate level of care for each patient. From these analyses many aspects of trauma care can be improved and optimized, such as determining the appropriate amount of ICU beds and other trauma care resources @tqip.

It is tempting to apply existing models of TQIP assessments on Indian trauma data to identify causes of the observed disparities in trauma care performance in HIC and LMIC, and propose measures to harmonize performance. Unfortunately as TQIP often originate in high-income countries (HIC), they are either not applicable nor easily transferred to the hospital conditions of LMIC @frailty @gunshot  @onemilliondead. Applying TQIPs outside of their original context or even within a subset of its own origin is difficult even without taking income bracket in account, for instance firearm violence is a considerable bigger portion of trauma patients in some US contexts, than the national average @gunshot. For trauma patients, there exist significant structural differences between HIC and LMIC. These differences are in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead. One example of this is the differences in helmet use for motorcyclists and their passengers, as well as differences of displacement in vehicle fleets in LMIC versus HIC. These factors have tremendous impact on the type and severity of injuries patients in the ER have, and subsequently on their care and outcome. These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.

Kommentar på Mgerdin kommentar: jag snodde din definition här. Jag håller med om att detta inte är den mest relevanta informationen men för "kontrollmekanism"-arbetet som ska in på måndag här har KI en checklista och de vill ha 3-5 sidor bakgrund trots att vi jobbar med max 1-1.5 så jag måste feta ut. Detsamma för kommentaren i Aim jag är medveten om att det är felplacerat eller irrelevant.

Knowlege gap
------------
There are compelling reasons to believe there is a shortage of ICU beds in India, but the scope of this shortage and the potential for reduction of trauma mortality if this shortage would be remedied is unknown.

WHO and local health care workers believe there is a shortage of ICU beds in Indian hospitals, and that the lives of many patients would have been salvageable if the appropriate ICU resources had been available. The differences of in-hospital trauma mortality in Indian hospitals was double that of high-income countriesin, and part of this difference could be explained by a shortage of ICU beds in India @indiantrauma

@@

@@

Propensity score matching in itself introduces a source of bias from unfit assumptions. The terminology and potential sources of bias are explained chapter Study Design. In choosing predictor variables the researcher could introduce an overdetermined set and a self-fulfilling prophecy. For this reason the dataset shall be randomly split into two parts of 40% and 60% of the patients. The 60% shall be used to produce the propensity score matching, and the remaining 40% shall be used to verify it.

@@

@@

Second step is to identify predictor variables: covariates most predictive of the dependent variable, as described in chapter Predictor selection. In this context it means to first identify the conditions and medical interventions received that correlate best to being admitted to ICU care. This selection is based on clinical experience as well as analysis of the Training Data. In the data analysis all properties of the patients that were sent to the ICU will be compared to the patents who did not, in order to identify properties that distinguishes this cohort., Familiarity with the collection of source data is crucial to minimize the risk to choose irrelevant predictors and introduce error.

@@

Fourth step is PSM. In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertedly vastly under- or overestimate the  need of ICU beds remain.

The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one trom Training Data.

@@

All patients with incomplete observations for any predictors will be discarded. For relevant qualitative variables, missing data did not automatically warrant dismissal, if the missing data could be indirectly proven. For example if the quantitative yes/no for an examination like a CT scan was missing, it was deemed a 'yes' if there was a CT scan result recorded. If both were missing, the patient data was dismissed.
All vital parameters are considered mandatory, irregardless of patient condition so patients without data of heart rate, systolic blood pressure or respiration rate at any time of their hospital stay, have been discarded.  
FRÅGA: right?

@@

FRÅGA: det finns pat som är intuberade "before arrival" men som inte är noterade som transferred från ett annat sjukhus. De har kommit med motor rickshaw. Är det rimligt att anta att de är transferred från ett sjukhus iallafall för vem har annars intuberat dessa patienter?

<!-- 
The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, where only a small proportion of patients had been taken within the first hour but a substantial proportion of patients had been taken to the operation theatre after 24 hours. There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity. 
-->
The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, and operations performed in the first hour tend to be more acute than operations performed within the first day of hospitalization. There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity. 
FRÅGA: borde man ta operation within 24 hours och komplettera med text based operationstyp?

>>>>

Trauma globally and in India
----------------------------
Trauma is a substantial global health problem that disproportionately affects low and middle-income countries (LMIC), resulting in significant mortality and morbidity @traumaLMIC. Despite having only 1% of the global vehicle fleet, a disproportionate 10% of global traffic casualties occur in India @indiantrauma. The annual mortality from injury in India amount to one million casualties @onemilliondead. Half of trauma mortality in India occur in-hospital. There is no national trauma registry in India, but isolated studies have shown that in-hospital mortality within 30 days of hospitalization is higher than similarly injured patients in HIC hospitals @indiantrauma.

@@

Health care professionals in many LMIC testify that their medical assessment oftentimes is that a trauma patient should be admitted to the ICU, but are not because such resources are scarce, overloaded and unavailable. Their medical judgement is that the lives of these patients were salvageable, and that these patients died as a direct result of not being admitted to the ICU @stuffLMICdoctorssay. Globally, an estimated two million lives could be saved annually, if the appropriate and timely trauma care provided by hospitals in many high-income countries were available and provided in low to-middle-income countries (LMICs) @indiantrauma

A diverse mix of public, for-profit and not-for-profit private actors provide health services in India, where private providers dominate. According to a WHO 2022 report, India has 76 709 non-ventilator supported and 39 476 ventilator-supported ICU beds @ICUbeds.

<!-- KommentarJohanna: I suggest adding more details about the context in India, the health care system, how common trauma is in India, how many ICU beds there are, why ICU care is important for the trauma population and so on.. 
-->
<!-- KommentarMartin:You can definitely write more about the health care system in India, the trauma system, the availability of ICUs, how common ICU care is in trauma, why trauma patients need ICU care, how long they stay, outcomes etc -->

<!-- Respons: Datan som står att finna om ICU vård resurserna i Indien är lite spretig, men enligt WHO så är den runt 8 per 100.000 och andra trovärdiga bedömare är där nånstans +- 50%. Det är i paritet med Sverige? Förklaringen står kanske att finna i att definitionen av en ICU bed inte är vedertagen men trots mycket läsande innantill har jag inte kunnat avfärda eller förklara detta. 
Och det lyfter frågan: det finns lite olika levels av ICU beds, vilken är det vi menar? Eftersom statistiken på ICU resurser är dåligt differentierad och TITCO datasetet inte specificerar det, ska frågan undvikas för att inte öka komplexiteten för sig själv? -->

@@

TQIP often originate in high-income countries (HIC), they are either not applicable nor easily transferred to the hospital conditions of LMIC @frailty @gunshot  @onemilliondead. For trauma patients, there exist significant structural differences between HIC and LMIC. These differences are in factors such as injury mechanisms and severity, health care resources as well as patient make-up @frailty @indiantrauma @traumaLMIC @onemilliondead. These differences make a direct transferral of HIC TQIP models onto LMIC irrelevant or misleading and there is a need to develop TQIP assessment applicable to local conditions.

<!-- Ni vil båda flytta fokus från detta, och jag förstår inte riktigt varför det är med kopplingen är luddig. Är det för att säga: det finns TQIPs som ger en fingervisning om hur ICU ska dimensioneras, bestyckas, bemannas och bedrivas. Men det finns inga färdiga TQIPs som kan användas för att kvantifiera ICU-bristen i Indien och vi tror det är vanskligt att använda de befintliga eftersom de skapats under så olika förhållanden. Och därför måste vi börja från noll? -->

Knowlege gap
------------
There are compelling reasons to believe there is a shortage of ICU beds in India, but the scope of this shortage as well as the potential for reduction of trauma mortality if this shortage would be remedied is unknown.

WHO and local health care workers believe there is a shortage of ICU beds in Indian hospitals, and that the lives of many patients would have been salvageable if the appropriate ICU resources had been available @ICUbeds. The differences of in-hospital trauma mortality in Indian hospitals was double that of high-income countries, and part of this difference could be explained by a shortage of ICU beds in India @indiantrauma .

@@

<!-- Kommentar: inför datatvätten, hur vill vi göra med  småbarnen? -->

@@

<!--  
Propensity score matching in itself introduces a source of bias from unfit assumptions. The terminology and potential sources of bias are explained chapter Study Design. In choosing predictor variables the researcher could introduce an overdetermined set and a self-fulfilling prophecy. For this reason the dataset shall be randomly split into two parts of 40% and 60% of the patients. The 60% shall be used to produce the propensity score matching, and the remaining 40% shall be used to verify it. 
-->

<!-- Fråga: vad är distinktionen mellan bias och error med PSM egentligen? är det error eller bias som uppstår om jag ex skulle välja ICU-sjukhuset som prediktor? Är det bias att ha för generösa calipers? Som jag förstår finns risk att få vilket resultat som helst med val metod för matching och limitations ? Och att jag väljer calipers som producerawr resultat som jag tycker känns "rimliga"? Ex jag kanske väntar mig att ICU bristen är 20% för jag föreställer mig att de patienter jag söker är de som är "på ribban" alltså de matchar de "friskaste" 20% av dem som var på ICU. Det känns mindre sannolikt att en patient inte fått ICU om den är jämförbar med någon som var en av de mest skadade 20% på ICU.   -->

@@

<!-- ÅTERKOM inte klart bara gissat -->

@@

Second step is to identify predictor variables: covariates most predictive of the dependent variable, as described in chapter Predictor selection. In this context it means to first identify the conditions and medical interventions received that correlate best to being admitted to ICU care. This selection is based on clinical experience as well as analysis of the Training Data. In the data analysis all properties of the patients that were sent to the ICU will be compared to the patents who did not, in order to identify properties that distinguishes this cohort. Familiarity with the collection of source data is crucial to minimize the risk to choose irrelevant predictors and introduce error.

@@

Fourth step is PSM. In PSM members of the subset that did not get ICU care but had similar predictors - and therefore similar propensity scores - as the subgroup that did get ICU care, are matched to patients who did go to the ICU. There are a few different ways to perform this matching but they are all imperfect because propensity scores are unique and perfect matching is unattainable There are control mechanisms and limitations but a risk of applying too fix or too loose matching criteria the outcome could inadvertedly vastly under- or overestimate the  need of ICU beds remain.  
The same procedure is repeated on the Validation Data using the same predictors and the same matching criteria, and this should produce a cohort comparable to the one trom Training Data. The proportion of this subset versus the subset that did receive ICU care is also the proprtion of ICU bed shortage. If, for every patient that did receive ICU care, there was another patient with similar predictors that wasn’t admitted due to scarcity of ICU resources, that would mean the ICU resources are 50% of what they ought to be.

@@

All patients with incomplete observations for any predictors will be discarded. For relevant qualitative variables, missing data did not automatically warrant dismissal, if the missing data could be indirectly proven. For example if the quantitative yes/no for an examination like a CT scan was missing, it was deemed a 'yes' if there was a CT scan result recorded. If both were missing, the patient data was dismissed.
All vital parameters are considered mandatory, irregardless of patient condition so patients without data of heart rate, systolic blood pressure or respiration rate at any time of their hospital stay, have been discarded.

@@

<!--
FRÅGA datatvätt: det finns pat som är intuberade "before arrival" men som inte är noterade som transferred från ett annat sjukhus. De har kommit med motor rickshaw. Är det rimligt att anta att de är transferred från ett sjukhus iallafall för vem har annars intuberat dessa patienter?
-->

<!-- 
The parameter "taken to the operation theatre" is a heterogenous qualitative parameter, where only a small proportion of patients had been taken within the first hour but a substantial proportion of patients had been taken to the operation theatre after 24 hours and many of these operations are lower priority orthopedic operations with no correlation to ICU admission. There is a need to consider the free text field operation type to filter out for operation types with higher propensity for ICU admission from operations with lower propensity. 
FRÅGA prediktorval: borde man ta operation within 24 hours och komplettera med text based operationstyp?
-->
